Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2017-09-25 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
paper2 pre-sub tweaks

_must be 350 words, currently 348_

@@

We provide a concise, accessible description of a flexible model of the evolution of insecticide resistance. We use the model to develop a mechanistic picture the evolution of insecticide resistance and how it is likely to respond to potential insecticide use strategies.  The aim is to reach an audience unlikely to read a more detailed modelling paper. The model itself, as described here, represents two independent genes coding for resistance to two insecticides. This allows us to represent the use of insecticides in isolation, sequence and mixtures.

##### Results
We use the model to demonstrate the evolution of resistance under different inputs and how this fits with intuitive reasoning about selection pressure. We show that using an insecticide in a mixture, relative to alone, always prompts slower evolution of resistance to that insecticide. However, when the resistance to both insecticides is considered, resistance thresholds may be reached later for a sequence relative to a mixture. Increasing the ability of insecticides to kill susceptible mosquitoes (effectiveness), has the most influence on favouring a mixture over a sequence because one highly effective insecticide provides more protection to another in a mixture.

@@

## Background
Insecticide resistance is a problem for malaria [@WHO2012][@Ranson2016][@Hemingway2016] other vector borne diseases [@IRAC2011] and agriculture [@FAO2012]. Malaria alone still results in hundreds of thousands of deaths per year. Recent malaria control efforts have centred on treated bed nets and indoor residual spraying, both reliant on insecticides. Treated nets were recently estimated to contribute 68% and indoor residual spraying 13% to averting more than 500 million falciparum malaria cases between 2000 and 2015 [@Bhatt2015]. A recent malaria transmission model [@Churcher2016] predicts that even low pyrethroid resistance is likely to increase malaria incidence in Africa by reducing the performance of bed nets.

The WHO produced a Global Plan for Insecticide Resistance Management in malaria vectors (GPIRM)[@WHO2012] which includes recommendations on operational strategies for managing resistance including the use of insecticide mixtures when they become available.  Efforts are under way to develop new insecticides that will be effective in the light of existing resistance and allow additional options within insecticide resistance management. The Innovative Vector Control Consortium (IVCC) was set up in 2005 to work with industry in the development of new vector control tools and particularly new insecticides with new modes of action to address insecticide resistance in malaria transmitting mosquitoes [@Hemingway2006][@IVCC2016]. Three new insecticides are now in development [@IVCC2016] and likely to be available after 2020 [@Ranson2016]. It is important that decisions about how best to use the new insecticides, with a clear mindset of delaying the evolution of resistance, are made before insecticides are released [@Hemingway2016].

A recent comprehensive review of strategies to avoid resistance evolution across pesticides and drugs [@Consortium2013] concluded that mixtures (combination of molecules) are usually the best resistance management strategy. This was based on both empirical and modelling work. Modelling studies have investigated the evolution of insecticide resistance in insecticide mixtures including in a public health context e.g. [@Curtis1985][@Mani1985][@Roush1989] but much of the work was done more than 20 years ago and there remained some confusion about the results [@Levick2017]. In an earlier paper [@Levick2017] we described the technical details of a flexible model used to investigate the relative benefits of mixtures and sequences. Here we provide an accessible summary of the model and use selected parameter values to describe mechanistically how the evolution of resistance can be expected to change in different scenarios. This mechanistic understanding can contribute to the debate on the relative merits of different insecticide strategies extending existing frameworks [@Roush1989][@IRAC2011][@FAO2012][@WHO2012][@Consortium2013].

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. We assume both that genes conferring resistance are already present in the population and that resistance to each insecticide is coded by a single gene. These assumptions are consistent with previous modelling work. In support of the first assumption insecticide resistance has been termed 'pre-adaptive' meaning that resistance alleles are present in very low numbers even prior to exposure to novel insecticides [@Ffrench-Constant2013][@Liu2015] as has been documented in blowflies [@Hartley2006]. In support of using a single gene representation of resistance evidence suggests that whilst polygenic resistance is common, control failures, particularly those due to target site resistance, are mostly due to single major genes [@Ffrench-Constant2013].

## Methods
The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One gene or locus is represented for each insecticide. Each individual has two copies of the gene and there are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Thus individuals can either have both resistance genes (RR homozygous), both susceptible genes (SS) or one of each (SR, heterozygous). The combination of the two genes is termed the genotype. Fitness is the main currency of the model, representing how much each genotype survives and reproduces. The fitness of the SS genotype in absence of the insecticide is set at a maximum value of 1. All other fitnesses are less, determined by alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide, the dominance of the resistance allele and the ability of resistance to restore fitness in the presence of the insecticide (see Table 1).

@@

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 1. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1 (equivalent to the unexposed SS), a value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. The 'selection coefficient' of resistance, that may be seen referred to elsewhere [@Curtis1985], can be calculated by multiplying resistance restoration by insecticide effectiveness. Cost of resistance determines the fitness of RR in the portion of the population not exposed to the insecticide.

@@

The simulation proceeds through generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours but we do not explore that here.

@@

The model is implemented in R [@RCoreTeam2016], the code is hosted on Github [@South2017] and we provide an on-line user interface [@South2017a] (Fig 2) enabling the reader to change inputs and run the model themselves.

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 3 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 4A) through the even simpler mechanism that less change is required to reach the threshold.

@@

#### Caveats
There is much evidence of the involvement of multiple genes in resistance mechanisms. Our model, in common with most previous ones, assumes a single ‘resistance’ gene per insecticide. In most cases we believe this is reasonable. Target-site resistance has been shown to be mostly determined by single genes, and control failures in general have mostly been attributed to the effect of single major genes [@Roush1989][@Ffrench-Constant2004][@Ffrench-Constant2013]. Control doses of insecticides in the field are likely to lead to the selection of single major genes whereas the lower doses used to select laboratory strains are more likely to lead to polygenic resistance [@Ffrench-Constant2013]. To represent polygenic resistance a different 'quantitative genetic' approach would be required [@Via1986][@Firko1990]. We are only aware of preliminary published work applying a quantitative genetic approach to assessing insecticide use strategies in the context of polygenic resistance [@Via1986][@Gardner1998]. One suggestion is that evolution of resistance to one insecticide in a mixture could be associated with a decrease in tolerance to the other [@Via1986]. If it becomes clearer that polygenic resistance is an issue then we would welcome a comparison of the results presented here to those of quantitative genetic models.

Even if the assumption that single genes are responsible for control failures is correct there is still considerable other uncertainty around the development of insecticide resistance in operational field settings [@Ranson2016]. The model we present here is able to cover much of the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). There is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences.

In addition we provide an on-line user interface to the model (Fig 2) [@South2017a] enabling readers to investigate model behaviour in areas of parameter space we have not shown. The user-interface allows inputs to be set for two scenarios and results to be viewed side by side. For example you can set a higher insecticide effectivenesses in one scenario to see how this favours mixtures over sequences (Fig 2) or increase the exposure levels to see how this favours sequences.

There are other model behaviours that can be changed by modifying input files. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying input files would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately.

@@

##### Funding
This work was supported by: Bill and Melinda Gates Foundation, grant #37999.01 to IMH; and Wellcome Trust Institutional Strategic Strengthening fund grant #105620/Z/14/Z to Liverpool School of Tropical Medicine; sub-funding to IMH; and a small grant from Bayer S.A.S. to AS.

>>>>

@@

We provide a concise, accessible description of a flexible model of the evolution of insecticide resistance. We use the model to develop a mechanistic picture of the evolution of insecticide resistance and how it is likely to respond to potential insecticide use strategies.  The aim is to reach an audience unlikely to read a more detailed modelling paper. The model itself, as described here, represents two independent genes coding for resistance to two insecticides. This allows us to represent the use of insecticides in isolation, sequence and mixtures.

##### Results
We use the model to demonstrate the evolution of resistance under different scenarios and how this fits with intuitive reasoning about selection pressure. We show that using an insecticide in a mixture, relative to alone, always prompts slower evolution of resistance to that insecticide. However, when resistance to both insecticides is considered, resistance thresholds may be reached later for a sequence relative to a mixture. Increasing the ability of insecticides to kill susceptible mosquitoes (effectiveness), has the most influence on favouring a mixture over a sequence because one highly effective insecticide provides more protection to another in a mixture.

@@

## Background
Insecticide resistance is a problem for malaria control[@WHO2012][@Ranson2016][@Hemingway2016] other vector borne diseases [@IRAC2011] and agriculture [@FAO2012]. Malaria alone still results in hundreds of thousands of deaths per year. Recent malaria control efforts have centred on treated bed nets and indoor residual spraying, both reliant on insecticides. Treated nets were recently estimated to contribute 68% and indoor residual spraying 13% to averting more than 500 million falciparum malaria cases between 2000 and 2015 [@Bhatt2015]. A recent malaria transmission model [@Churcher2016] predicts that even low pyrethroid resistance is likely to increase malaria incidence in Africa by reducing the performance of bed nets.

The WHO produced a Global Plan for Insecticide Resistance Management in malaria vectors (GPIRM)[@WHO2012] which includes recommendations on operational strategies for managing resistance including the use of insecticide mixtures when they become available.  Efforts are under way to develop new insecticides that will be effective in the light of existing resistance and allow additional options within insecticide resistance management. The Innovative Vector Control Consortium (IVCC) was set up in 2005 to work with industry to develop new vector control tools and particularly new insecticides with new modes of action to address insecticide resistance in disease transmitting mosquitoes [@Hemingway2006][@IVCC2016]. Three new insecticides are now in development [@IVCC2016] and likely to be available after 2020 [@Ranson2016]. It is important that decisions about how best to use the new insecticides, with a clear mindset of delaying the evolution of resistance, are made before insecticides are released [@Hemingway2016].

A recent comprehensive review of strategies to avoid resistance evolution across pesticides and drugs [@Consortium2013] concluded that mixtures (combination of molecules) are usually the best resistance management strategy. This was based on both empirical and modelling work. Modelling studies have investigated the evolution of insecticide resistance in insecticide mixtures including in a public health context e.g. [@Curtis1985][@Mani1985][@Roush1989] but much of the work was done more than 25 years ago and there remained some confusion about the results [@Levick2017]. In an earlier paper [@Levick2017] we described the technical details of a flexible model used to investigate the relative benefits of mixtures and sequences. For that paper we investigated thousands of scenarios to explore potential outcomes. In contrast here we provide an accessible summary of the model and use selected parameter values to describe mechanistically how the evolution of resistance can be expected to respond. This mechanistic understanding can contribute to the debate on the relative merits of different insecticide strategies and extend existing frameworks [@Roush1989][@IRAC2011][@FAO2012][@WHO2012][@Consortium2013].

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. We assume that genes conferring resistance are already present in the population and that resistance to each insecticide is coded by a single gene. These assumptions are consistent with previous modelling work [@Consortium2013]. In support of the first assumption insecticide resistance has been termed 'pre-adaptive' meaning that resistance alleles are present in very low numbers, most likely as a mutation-selection balance, even prior to exposure to novel insecticides [@Ffrench-Constant2013][@Liu2015] as has been documented in blowflies [@Hartley2006]. In support of using a single gene representation of resistance evidence suggests that whilst polygenic resistance is common, control failures, particularly those due to target site resistance, are mostly due to single major genes [@Ffrench-Constant2013].

## Methods
The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One gene or locus is represented for each insecticide. Each individual has two copies of the gene and there are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Thus individuals can either have both resistance genes (RR homozygous), both susceptible genes (SS) or one of each (SR, heterozygous). The combination of the two genes is termed the genotype. Fitness is the main currency of the model, representing how much each genotype survives and reproduces. The fitness of the SS genotype in absence of the insecticide is used as the reference genotype and assigned a value of 1. All other fitnesses are assumed to be less than 1, determined by alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide against SS genotypes, the dominance of the resistance allele and the ability of resistance to restore fitness of the RR genotype in the presence of the insecticide (see Table 1).

@@

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 1. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1, equivalent to the unexposed SS and meaning the RR genotype is completely unaffected by the insecticide. A value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. Cost of resistance determines the fitness of RR in the portion of the population not exposed to the insecticide.

@@

The simulation proceeds through non-overlapping generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness, the latter depending on their absolute fitness (Fig 1) and level of exposure. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours but we do not explore that parameter here.

@@

A major aim of this paper is to us modern techniques to make the modelling accessible to a wider community. The model is implemented in R [@RCoreTeam2016], the code is open-source and hosted on Github [@South2017] including the code to generate both the figures and text of this paper. We provide an on-line user interface [@South2017a] (Fig 2) accessing the same code and enabling readers with no coding experience to change inputs and run the model themselves.

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 3 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 4A) through the even simpler mechanism that a smaller cahnge in frequency is required to reach the threshold.

@@

#### Caveats
There is evidence of the involvement of multiple genes in resistance mechanisms[@Ffrench-Constant2004][@Ffrench-Constant2013]. Our model, in common with almost all previous ones [@Bourguet2010], assumes a single ‘resistance’ gene per insecticide. In most cases we believe this is reasonable. Target-site resistance has been shown to be mostly determined by single genes, and control failures in general have mostly been attributed to the effect of single major genes [@Roush1989][@Ffrench-Constant2004][@Ffrench-Constant2013]. Control doses of insecticides in the field are likely to lead to the selection of single major genes whereas the lower doses used to select laboratory strains are more likely to lead to polygenic resistance [@Ffrench-Constant2013]. To represent polygenic resistance a different 'quantitative genetic' approach would be required [@Via1986][@Firko1990]. A recent review of 187 models of the evolution of resistance [@Bourguet2010] did not find any that represented quantitative multiple gene resistance. We are only aware of preliminary published work applying a quantitative genetic approach to assessing insecticide use strategies in the context of polygenic resistance [@Via1986][@Gardner1998]. If it becomes clearer that in some cases polygenic resistance is more of an issue than previously thought then there will be a need for appropriate quantitative genetic models. We would be keen to contribute to that development.

Even if the assumption that single genes are responsible for control failures is correct there is still considerable other uncertainty around the development of insecticide resistance in operational field settings [@Ranson2016]. Our model is able to represent much of that uncertainty, yet in this paper we restrict our explorations to a limited region of parameter space. Here we hold most inputs constant at intermediate values while modifying single inputs in isolation. This is a deliberate tactic to develop understanding and communicate the key mechanisms. There is the potential that the system may behave differently in the field. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences.

In addition we provide an on-line user interface to the model (Fig 2) [@South2017a] enabling readers with no coding experience to investigate model behaviour in areas of parameter space we have not shown. The user-interface allows inputs to be set for two scenarios and results to be viewed side by side. For example you can set a higher insecticide effectivenesses in one scenario to see how this favours mixtures over sequences (Fig 2) or increase the exposure levels to see how this favours sequences.

There are other model behaviours that can be changed by modifying model inputs within the computer code. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying model inputs would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately.

@@

##### Funding
This work was supported by: Bill and Melinda Gates Foundation, grant #37999.01 to IMH; and Wellcome Trust Institutional Strategic Strengthening fund grant #105620/Z/14/Z to Liverpool School of Tropical Medicine; sub-funding to IMH; a small grant from Bayer S.A.S. to AS; and IVCC funding to IMH.
